---
title: "Updating Documentation"
description: "Steps for updating the web-based documentation/help docs"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Updating Documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
---

```{r, eval=TRUE, echo=FALSE, include=FALSE}

code_reponame_with_owner <- EJAM:::repo_from_desc()           # "OWNER/REPONAME"
code_reponame <- gsub(".*/", "", code_reponame_with_owner)    #       "REPONAME"
code_reponame_url <- EJAM:::repo_from_desc(get_full_url = T)   
# "https://github.com/OWNER/REPONAME"

docs_reponame_with_owner <- EJAM:::repo_from_desc('github.io')# "OWNER/REPONAME"
docs_reponame <- gsub(".*/", "", docs_reponame_with_owner)    #       "REPONAME"
docs_reponame_url <- EJAM:::repo_from_desc('github.io', get_full_url = T) 
# "https://OWNER.github.io/REPONAME"

data_reponame_with_owner <- desc::desc(file = system.file(
  "DESCRIPTION", package = "EJAM"))$get("ejam_data_repo")     # "OWNER/REPONAME"
data_reponame <- gsub(".*/", "", data_reponame_with_owner)    #       "REPONAME"
data_reponame_url <- paste0("https://github.com/", data_reponame_with_owner) 
# "https://github.com/OWNER/REPONAME"
```

This document addresses how to update the documentation supporting the package.

## Updating package documentation

The R package is documented in webpages (and in standard help documents viewable
in RStudio). Documentation will generally need to be updated in preparation for
a new release of the package. Updates may be needed in the following:

1.  [Package: overall, package-wide information/ documents]

2.  [Functions: reference docs explaining the R functions]

3.  [Datasets: reference docs explaining the datasets] (in EJAM/R/data\_\*.R
    files)

4.  [Articles: vignettes helping R users with the package] (in
    `EJAM/vignettes/*.Rmd`)

5.  [User Guide: help for people using the web app] (in `EJAM/inst/app/www/` as
    a .pdf file)

6.  [Spell-checking package-wide]

7.  [Website: web pages providing all of the above in html format], via the
    `{pkgdown}` package

Each is explained below:

### Package: overall, package-wide information/ documents

-   Package-wide documentation files can be updated/edited and include the
    following: `DESCRIPTION`, `README.Rmd` (for README, **edit the
    README[.Rmd]{.underline}*NOT* the README.[md]{.underline} file**),
    R/EJAM-package.R (for `?EJAM-package`),
    `NEWS.md` used to make [NEWS](../news/index.html), 
    `CONTRIBUTING.md` used to make [CONTRIBUTING](../CONTRIBUTING.html),
    [LICENSE](../LICENSE.html) based on `inst/LICENSE.md`,
    [CITATION](../authors.html#citation) based on `inst/CITATION`, 
    and others.

-   Some of these are used by and shown by the github.com repository, notably
    the `README.md` that is generated from the `README.Rmd` file and appears on
    the github repository as
    [README](%60r%20paste0(code_reponame_url,%20%22?tab=readme-ov-file#readme%22)%60)
    and in the pkgdown site as [README](../index.html), and the `DESCRIPTION`
    file that stores package version number and URLs of the package repository,
    etc.

### Functions: reference docs explaining the R functions

-   `{roxygen2}` package tags define the documentation in the R/\*.R files that
    contain the source code and documentation of functions and datasets.

-   While many functions or objects have their own .R files, some .R files
    contain code and documentation for multiple functions, so not every function
    has a .R file of its own.

-   some utility/helper functions, mostly not exported, are stored in files
    prefixed with "utils\_" such as `R/utils_indexblocks.R` documenting `indexblocks()`
    (but not all helper functions are in files named that way).

-   you might find it useful to check docs with `tools::checkDocFiles`("EJAM")

### Datasets: reference docs explaining the datasets

-   all datasets are documented in the files named with a prefix of "data\_"
    such as `R/data_blockgroupstats.R` which documents the `?blockgroupstats`
    dataset.

-   most dataset documentation is created/updated alongside the code that
    created or updates those datasets. That is best done starting with the
    overarching file called `data-raw/datacreate_0_UPDATE_ALL_DATASETS.R` which
    helps walk through updating various datasets as needed and their
    documentation is often updated in that same process.

-   Many `R/*.R` files for datasets are created automatically, not edited by
    hand, through the `dataset_documenter()` function as used in the scripts
    within `data-raw/datacreate_*.R` while others are updated by editing the .R
    file directly.

### Articles: vignettes helping R users with the package

-   Articles (called vignettes, here) are written/edited as .Rmd files in the `vignettes` folder. You can
    test one using something like `pkgdown::build_article()` before trying to
    rebuild the whole pkgdown site.

-   New articles or renamed files require edits to the `_pkgdown.yml` file also.

### User Guide: help for people using the web app

-   If you change the UI then new screenshots are needed for a new version of
    the User Guide that was stored as a .pdf document.

### Spell-checking package-wide

-   To run a spell check (see terms flagged as possible problems)

```{r, eval=FALSE}
    x <- spelling::spell_check_package()
    x <- data.frame(
      frq = sapply(x$found, function(z) {length(unlist(z))}), 
      word = x$word)
    x <- x[order(x$frq), ]
    rownames(x) <- NULL
    x
```

-   Then you can use ctrl-shift-F to search in all files for a flagged term to
    check if it is actually a problem

-   If you wanted to update the WORDLIST (list of words to ignore spelling of)

```{r, eval=FALSE}
    # 1. Add obvious dataset names documented to the WORDLIST
    datanames = gsub('\\.R$', '', gsub('^data_', '', dir('./R', pattern = '^data_.*R')))
    datanames = datanames[!grepl('aaaaaaaaaaaaa|xxxxxxxxxx', datanames)]
    # 2. Add names of functions to the WORDLIST
    funcnames = EJAM:::functions_in_pkg( 'EJAM' )$object
    wordlist = readLines('inst/WORDLIST')
    wordlist = sort(unique(union(wordlist, c(datanames, funcnames))))
    # Write the revised list of words to the file
    writeLines(wordlist, 'inst/WORDLIST')
```

### Website: web pages providing all of the above in html format

The web-based documentation pages should be updated right after any of the above
are edited/created. These pkgdown-based webpages should be updated by someone
who is managing the package, as follows:

-   The first step in updating those web pages is updating the yml file. The
    web-based documentation pages are organized by a file called `_pkgdown.yml`
    -- see `usethis::edit_pkgdown_config()` -- which specifies the contents of
    the webpages of documentation, such as the names of any vignettes files to
    be shown in the list of articles, which functions are included in the
    reference pages index, etc. Any new or removed or renamed functions/
    vignettes/ datasets, or changes in what is exported, will require edits to
    the .yml file.

-   The pkgdown-based webpages should be updated by someone who is managing the
    package, and they should use the `update_pkgdown()` function found in
    `data-raw/datacreate_0_UPDATE_ALL_DOCUMENTATION_pkgdown.R` You have to
    `source()` that file and then use the function that it defines. Via that
    function, the pages get rebuilt by the `{pkgdown}` R package, based on the
    above files and all other documentation from .R and .md files. For more
    information about using the `{pkgdown}` package to create documentation in
    the form of webpages, see [pkgdown
    articles](https://pkgdown.r-lib.org/articles/pkgdown.html). The
    documentation is created/updated using the `{roxygen2}` and `{pkgdown}` R
    packages, which read information from the .R files in the R folder, from
    .Rmd file vignettes (aka "articles") in the vignettes folder, and from the
    `_pkgdown.yml` file in the root folder of the source package. The
    `{roxygen2}` package reads the .R files to create the .Rd files in the
    `man/` directory. The `{pkgdown}` package uses all those files as well as
    `DESCRIPTION` and others, to generate a website for the package, first
    created as files in the `docs/` folder, which are then published via github
    actions and the github repository settings, to be hosted on GitHub Pages.
    The packages `{devtools}` and `{usethis}` are also relevant.

-   the package's github repository settings define which branch and which
    folder should hold the html files for deployment (see github and pkgdown
    documentation on that)

-   the package's Github Actions can be used to deploy the webpages when
    triggered by an event like any push to the main branch, for example.

-   depending on which folder you publish/deploy from, you might have to remove
    that folder from `.gitignore` in case it was listed there via some
    `{pkgdown}` or `{usethis}` function.
