---
title: "Testing EJAM Functions with Unit Tests"
description: "How to run the suite of {testthat} unit tests"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Testing EJAM Functions with Unit Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
---

```{r developernote, eval=FALSE, echo=FALSE, include=FALSE}
#  *>>>>>>>>>> Developer note: vignettes need to be tested/edited/rebuilt regularly <<<<<<<<<<<*
#    - **See ?pkgdown::build_site** and script in EJAM/data-raw/- EJAM uses the pkgdown R package to build help and articles/ vignettes as web pages
```

```{r SETUP_default_eval_or_not, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.height = 6, fig.width = 7,
  comment = "#>"
)
knitr::opts_chunk$set(eval = FALSE)
# https://r-pkgs.org/vignettes.html
```

## Notes for developers, on writing or running unit tests in this package

### Checking test coverage (where more unit tests need to be written)

Some functions in the package have incomplete unit test coverage as of version
2.32.4 / mid 2025, so ideally more unit tests would be added in those cases.

You can see a grouped, organized list of unit test files *within the source
code* of a function test_interactively(), in
**`EJAM/tests/test_interactively.R`**

You should check which functions have adequate unit tests. One quick but
imperfect way is by using the rough draft function and notes found at
**`EJAM/tests/test_coverage_check.R`**. The test_interactively() function noted
above already prompts for using it, but you could separately `source()` it and
use it.

Note it is a bit tedious to figure out which functions have unit tests since
there is not a 1-to-1 relationship between functions like xyz() vs the
EJAM/R/xyz.R files defining functions vs EJAM/tests/testthat/test-xyz.R files
containing unit tests.

Also see the [covr package](https://covr.r-lib.org/) and
`covr::package_coverage()` and the related RStudio addin.

### Adding unit tests (for new functions or where unit tests not yet written)

Unit tests should be written when the occur:

-   where an existing function has no unit tests

-   where an existing function has only limited tests, and e.g., some
    combinations of parameter settings have not been tested in unit tests

-   while a new function is being written

See <https://testthat.r-lib.org> and `usethis::use_test()` for information on
how to add unit tests. The unit tests are in the `EJAM/tests` and
`EJAM/tests/testthat` folders.

#### Updating unit tests (when functions are updated)

Unit tests should be changed if necessary in these situations:

-   while an existing function is being modified, such as when the function's
    expected outputs/ warnings/ errors change.

-   if the web app UI changes - then special tests of the web app functionality
    can be created as explained in [Testing EJAM App with
    shinytest2](dev_run-shinytests.html). Also note that any screenshots for any
    [User Guide]{.underline} would need to be updated if the UI changes.

### Running unit tests

There are numerous unit tests in the `EJAM/tests/testthat/` folder that you can
run and add to as needed. They can be run automatically by the
`devtools::test()` function.

However, instead of running all the unit tests (run in alphabetical order by
default), you might want to test only some groups of related functions that you
are working on, and see a summary of which tests in which files from which
groups are passing, skipped, etc. To do this, note there is a file called
**`EJAM/tests/test_interactively.R`** that you can `source()` and then you can
use that function to run the unit tests interactively.

Assuming you want to use that function, whenever you add or rename or delete
unit test files in the `EJAM/tests/testthat/` folder you need to edit the list
of test file names in the source code of that function, test_interactively() or
you can try to use it and it will stop and tell you which files are missing from
the list in the source code in `EJAM/tests/test_interactively.R`

### Running tests of the web app functionality

There are recorded interactions used as tests of the web app, as explained in
[Testing EJAM App with shinytest2](dev_run-shinytests.html).

### Running unit tests via github workflow

There is also a github workflow in the package that runs unit tests,
automatically triggered by certain events, such as a pull request to the main or
development branch. See `EJAM/.github/workflows/run-test.yaml`

### Other notes

Be aware of whether the tests are using the last-installed version of the
package or the latest, possibly recently-edited, local source package version
that `devtools::load_all()` may be have loaded.

Some functions relied on the EJScreen API and if the API is not available, those
tests will be skipped or would fail.
