---
title: "Comparing Distances (Choosing a Radius)"
description: "Comparing Distances (Choosing a Radius)"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Comparing Distances (Choosing a Radius)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 80
---

```{r developernote, eval=FALSE, echo=FALSE, include=FALSE}
#  *>>>>>>>>>> Developer note: vignettes need to be tested/edited/rebuilt regularly <<<<<<<<<<<*
#    - **See ?pkgdown::build_site** and script in EJAM/data-raw/- EJAM uses the pkgdown R package to build help and articles/ vignettes as web pages
```

```{r SETUP_default_eval_or_not, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.height = 6, fig.width = 7,
  comment = "#>"
)
knitr::opts_chunk$set(eval = FALSE)
# https://r-pkgs.org/vignettes.html
```

```{r libraryEJAM, eval=TRUE, echo=FALSE, include=FALSE}
# rm(list = ls()); golem::detach_all_attached(); devtools::load_all()
 
if (!exists("blockgroupstats")) {library(EJAM)} # use installed version only if pkg not yet attached

dataload_dynamic('all') # varnames = all  currently means all defined by .arrow_ds_names

indexblocks()
```

Note: This article is a work in progress


# HOW TO ANALYZE PROXIMITY USING EJAM

An outline of how to use key functions is provided below. After these examples
is a discussion of background information and considerations in selecting
radius.

Also see the list of [functions related to comparing distances](../reference/index.html#comparing-distances-multiple-radius-values).

## RESIDENTIAL POPULATION GROUP PERCENTAGES BY DISTANCE AT BLOCK GROUP RESOLUTION

It is easiest to analyze distance increments based on each blockgroup's average
resident here. Block resolution is covered in a later section.

### WITHIN ONE RADIUS

#### Overall list of sites

At the *OVERALL LIST of sites* as a whole, which groups are *over-represented*
within X mile radius vs Statewide?

```{r ejamitrunnotshow, eval=TRUE, echo=FALSE, include=FALSE}
out <- ejamit(testpoints_100, radius = 3.1)
```

```{r ejamitshown, eval=FALSE, echo=TRUE, include=TRUE}
out <- ejamit(testpoints_100, radius = 3.1)
```

```{r, eval=TRUE, echo=TRUE}
ejam2ratios(out)
```

```{r barplot, eval=TRUE, echo=TRUE, fig.cap="Example of ejam2barplot() showing percent Asian among residents within 5 km of these 100 sites is more than two times the US rate overall", fig.height=5, fig.width=8}
ejam2barplot(out)
```

#### Just one site

At *JUST ONE SITE*, which groups are *over-represented* within X mile radius vs
Statewide?

```{r ejamitx, eval=TRUE, echo=FALSE, include=FALSE}
out1 <- ejamit(testpoints_100[2, ], radius = 3.1, quiet = TRUE)
```

```{r ejamitshown1pt, eval=FALSE, echo=TRUE, include=TRUE}
out1 <- ejamit(testpoints_100[2, ], radius = 3.1)
ejam2ratios(out1)
```

```{r onesiteyy, fig.cap = "Example of ejam2barplot() showing percent non-Hispanic White Alone among residents within 5 km of this one site is about 1.6 times the US rate overall", eval=TRUE, echo=TRUE}
ejam2barplot(out1)
```

#### Site by site comparison

Which groups are *over-represented* at *EACH SITE*, within X mile radius vs
Statewide

```{r tablesiteratios, eval=TRUE, echo=TRUE}
out <- testoutput_ejamit_10pts_1miles
x = round(data.frame(out$results_bysite)[, c("ratio.to.state.avg.pctlowinc", "ratio.to.state.avg.pctmin")], 2)
names(x) = fixcolnames(names(x),"r","shortlabel")
x = data.frame(sitenumber = 1:NROW(x), x)
x
```

Plot to compare sites, for just one residential population indicator

This plot shows that % low income among residents at sites 5 and 6 is more than
twice the relevant State average. It is near average at several other sites, and
is less than half the State average at sites 4 and 10.

```{r ejam2barplot_sites10, fig.cap="Example of ejam2barplot_sites()", fig.alt="Example of ejam2barplot_sites() showing to state average percent low income, one bar per site, where sites 5 and 6 have ratios above 2", eval=TRUE, echo=TRUE}
ejam2barplot_sites(out, "ratio.to.state.avg.pctlowinc", topn = 10, sortby = F)

## For raw values at key sites:
# ejam2barplot_sites(out, "pctlowinc")
```

### WITHIN MULTIPLE DISTANCES - COMPARING RADIUS CHOICES

#### Overall list of sites

At the *OVERALL LIST of sites* as a whole, which groups are *over-represented*
within X mile radius vs Statewide?

```{r ejamit_compare_distances, eval=TRUE, echo=TRUE}
radii <- c(1,2,3,10)
#radii <- c(1, 10) #  quicker example
pts <- testpoints_100[10:12, ]
```

See just the table

```{r ejamit_compare_distances_seecommandonly, eval=FALSE, echo=TRUE, include=TRUE}
x <- ejamit_compare_distances(pts, radii = radii, quiet = TRUE, plot = FALSE)
```

```{r ejamit_compare_distances_seeoutputonly, eval=TRUE, echo=FALSE}
x <- ejamit_compare_distances(pts, radii = radii, quiet = TRUE, plot = FALSE)
```

See the plot

```{r ejam2barplot_distances, fig.cap="Example of using ejam2barplot_distances()", fig.alt="Example of using ejam2barplot_distances() with distance on x axis and ratio to state average on y axis, with one line for each residential population group such as ratio to state average percent Hispanic, showing the indicator that increases the most as you get closer to site is Ratio to State avg % Hispanic", eval=TRUE, echo=TRUE, fig.height=5, fig.width=7}
# x <- ejamit_compare_distances(pts, radii = radii, quiet = TRUE) # in which default is plot=TRUE
# or 
ejam2barplot_distances(x)
```

--------------------------------------------------------------------------------

## RESIDENTIAL POPULATION GROUP DATA AT BLOCK GROUP RESOLUTION

Most of the EJAM functions use distance to the average resident of a block
group, which is calculated from the distance to each block's internal point and
uses the approximation that within a block the average resident and all
residents are as far as that internal point. For typical distances analyzed in
EJAM (e.g., 3 mile radius, or about 5 km) that is a good approximation, since
only about 2% of all US blocks are larger than 1 square mile.

If you need high spatial resolution (block by block) plots of an indicator as a
function of distance, you can directly work with getblocksnearby() or just use
the function plot_distance_by_pctd(). It uses the distance from the site to each
block's internal point (like a centroid) rather than just the distance to the
average resident in each block group.

### How residential population group percentages at *ONE SITE* vary as *a continuous function of distance*

Example of area where %Black is very high within 1 mile but drops by 3 miles
away

```{r plot_distance_by_pctd, fig.cap= "Example of using plot_distance_by_pctd()", fig.alt= "Example of using plot_distance_by_pctd() showing distance on x axis, indicator value within x miles on y axis, for %Black nonhispanic alone as function of distance at site 1, with striking decrease in percent from almost 80% around 0 miles away down to about 20% at 3 or 6 miles away, and showing 20% is the approx US 80th percentile (and state values overall shown as horizontal lines lower than that in this case)", eval=TRUE, echo=TRUE}
pts <- testpoints_100[3,]
y <- plot_distance_by_pctd(
  getblocksnearby(pts, radius = 10, quiet = T),
  score_colname = "pctnhba",
  sitenumber = 1)
#browseURL(url_ejscreen_report(lat = pts$lat, lon = pts$lon, radius = 0.5))
#browseURL(url_ejscreen_report(lat = pts$lat, lon = pts$lon, radius = 3))
```

Example of area that has higher %Hispanic as you go 10 to 30 miles away from
this specific point

```{r plot_distance_by_pctd2, fig.cap="Example of using plot_distance_by_pctd()", fig.alt="Example of using plot_distance_by_pctd(), showing an example where %Hispanic as a function of distance from site number 1 is very low compared to state or US overall, within any distance, but gradually rises from almost zero within 1 mile to just under 5% within about 30 miles away", eval=TRUE, echo=TRUE}
pts <- data.table::data.table(lat = 45.75464, lon = -94.36791)

y <- plot_distance_by_pctd(pts,
                      sitenumber = 1, score_colname = "pcthisp")
# browseURL(url_ejscreen_report(lat = pts$lat, lon = pts$lon, radius = 10))
# browseURL(url_ejscreen_report(lat = pts$lat, lon = pts$lon, radius = 30))
```

### *Step through all the sites* to see an indicator versus distance at each

Examples of sites analyzed here show some conclusions are very sensitive to the
radius used. The choice of radius in proximity analysis for some sites will lead
to a very different conclusion depending on the radius analyzed, if only a
single distance is checked or reported on. The relationship between distance X
and percent by residential population group within X miles can be positive, negative, or roughly
flat, etc., depending on the site and group. The residential population group percentage may be
above or below the US average or the State average within a given distance of
the site.

For the ten sites analyzed in this example, a wide range of patterns is found:

-   At site 5, % low income is extremely high very close to the site and falls
    sharply with distance but it remains quite high (still above 80th percentile
    of US or State) even within 4 miles.

-   At site number 2 here, % low income very close to the site is around the
    80th percentile in the State, and is around the US 80th percentile within
    about 1 mile, but then it falls to below State and then US average within
    around 2 and then 3 miles of the site.

-   At site 7, it is below average until about 8 miles, but is above US and
    State averages within 10 miles.

-   At site 9, it can be above or below average in State and/or in US, depending
    on the distance, but it is never as high as the 80th percentiles.

-   At sites 2, 3, 4, and 10, % low income is far below US and State averages
    within any distance shown here.

```{r step_thru_plots, eval=FALSE, echo=TRUE}
pts <- testpoints_10
s2b <- getblocksnearby(pts, radius = 10, quiet = T)
for (i in 1:NROW(pts)) {
  plot_distance_by_pctd(s2b, sitenumber = i, score_colname = "pctlowinc")
  readline() # hit any key to step through the plots
}
```

Block by block details are also easy to view in a map of all the nearby blocks,
as shown in the section on `plotblocksnearby()` and details of blocks near one
site.

### Cumulative Distribution plots of groups as *a continuous function of distance*

Out of all the residents within the area analyzed, see how some are mostly
nearby and others are further away, as a CDF plot. This shows the share of each
residential population group residing at various distances from sites, with distance from
nearest site on the x axis and the cumulative share of each group on the y axis
(of all residents within 10 miles, what percent have a site within X miles?). It
compares everyone nearby to just those who are among the percent low income, and
shows that, for example, a larger share of all the low income population within
10 miles actually live within about 6 miles than is the case for everyone within
10 miles. In other words, within the 10 mile radius circles, more of the low
income residents are closer to a site than are the non-low income residents or
all residents.

```{r plot_distance_cdf_by_group1, fig.cap="Example of using plot_distance_cdf_by_group()", eval=TRUE}
 # out <- ejamit(testpoints_10, radius = 10)
plot_distance_cdf_by_group(
  out$results_bybg_people,
  demogvarname = 'pctlowinc', demoglabel = 'Low Income')
```

--------------------------------------------------------------------------------

## MEAN DISTANCE BY RESIDENTIAL POPULATION GROUP

The analysis described above looks at residential population group percentages as a function of distance.
Another perspective is provided by looking at distance as a function of
residential population group. This means looking at the average distance or the whole
distribution of distances (or proximities) among all the residents within a
single residential population group, one group at a time, and comparing these groups.

### Overall list of sites

*Mean distance* of each group, at the *OVERALL LIST of sites* as a whole

To see a table of residential population indicators, showing the mean distance for each
group, compared to distance for those not in that residential population group:

```{r distance_mean_by_group_1, eval=TRUE}

out <- testoutput_ejamit_1000pts_1miles
## But try a larger radius to reveal more information:
# out <- ejamit(testpoints_100, radius = 10)

# see a table of demog indicators
distance_mean_by_group(out$results_bybg_people)

# for just 1 indicator
print(distance_mean_by_group(
  out$results_bybg_people, 
  demogvarname = 'pctlowinc', demoglabel = 'Low Income'))
```

To see a barplot, comparing just race/ethnicity groups:

```{r plot_distance_mean_by_group999, fig.cap="Example of using plot_distance_mean_by_group()", fig.alt="Example of using plot_distance_mean_by_group() showing 8 race ethnic subgroups on x axis and average distance for those in group as ratio to distance for residents not in given group, with yellow bars for hispanic and hawaiian/PI nonhispanic yellow meaning they are closer than overall average resident and orange for American Indian nonhispanic alone meaning they are only 93% as far as everyone else, 0.64 vs 0.69 miles away", eval=TRUE}
plot_distance_mean_by_group(out$results_bybg_people, 
                       demogvarname = names_d_subgroups,
                       demoglabel = fixcolnames(names_d_subgroups, "r", "shortlabel")
                       )
```

### Site by site comparison

*Mean distance* of each group, at *EACH SITE*, as ratio to mean of everyone else
nearby

Ratios at each site, of avg dist of group / avg dist of everyone else near site:

```{r distance_by_group_by_site, eval=FALSE, echo=TRUE}

out <- testoutput_ejamit_10pts_1miles
## But try a larger radius to reveal more information:
# out <- ejamit(testpoints_10, radius = 31)

x = distance_by_group_by_site(out$results_bybg_people)
x

# summary of closest group at each site and by how much
data.frame(site = colnames(x), 
           closestgroup = rownames(x)[sapply(x, which.min)], 
           their_avg_distance_as_pct_of_everyone_elses = round(100 * sapply(x, min, na.rm = TRUE), 0)
)
```

--------------------------------------------------------------------------------

## BACKGROUND AND OVERVIEW OF ISSUES IN PROXIMITY, DISTANCE, OR RADIUS

Distance from a potential source of environmental risk is often used as a simple
proxy for actual exposure or risk, when data are limited. Proximity analysis
uses distance (how far away) from a site, which is just the opposite of
proximity (how near) to a site.

Conclusions can be sensitive to the choice of radius, if only one radius is
reported on, as shown in [Step through all the sites to see an indicator versus
distance at each].

### Group's percentage at each distance versus distance for each population group

Two basic ways to report residential population percentages and risk are 1) showing residential population percentage as a
function or risk, and 2) showing risk as a function of residential population group:

1.  Residential population group percentage as a function of risk (or proximity): Many proximity analyses
    report percentage by distance or risk bin, such as % low income
    within 3 miles of a point. This expresses residential population shares as a function of
    proximity or risk. Sometimes other distance or risk bins are used, such as
    areas with risk above some cutoff. And sometimes instead of a continuous
    measure of percentage, the residential population data are used to categorize
    places in bins, such as areas in the top quartile of poverty rates.

2.  Risk (or proximity) as a function of residential population group: A different way to
    present this information is to report distance or risk as a function of
    residential population group -- this expresses distance within each residential population group,
    such as the average distance by group or the full distribution of risk
    within each group.

### Radius, radii, or continuous distance?

Proximity or distance as binary, categorical, or continuous metrics: Proximity
analysis has often relied on picking a single distance, a radius, and analyzing
conditions within that radius, such as all residents who live within 3 miles of
a point where a regulated facility is located. Sometimes an analysis will look
at two or even three distances. In some more sophisticated analyses, distance is
treated as a continuous measure. Some tools like EJScreen use a proximity metric
based on the inverse of distance (1/d) to provide a proximity score that gets
higher as distance gets smaller. But many EJ analyses still use a single
distance and analyze conditions within that distance.

EJAM makes it easier to do any of these types of analysis, because conclusions
can be sensitive to the choice of a single radius, and metrics and methods
provide different perspectives and reveal a richer picture of where people
actually live in relation to potential sources of exposure or risk.

### Comparisons within what distances or to what reference area(s)?

This is a tricky issue in proximity analysis: There is a subtle but vital
difference between proximity analysis using a single radius (binary distance)
and analysis using continuous distance. One way to think of this is that there
are two aspects of or degrees of proximity to consider when analyzing
residential population groups within a certain fixed distance (radius) from a single
facility point (or a whole set of facilities). These two ways of summarizing
proximity are complementary:

1.  Which groups tend to live nearby in the sense of being **within the radius
    versus outside the radius** selected? In other words, which groups are
    "over-represented" within X miles of the site? This treats proximity as a
    yes/no, binomial question -- **a resident is nearby or not**. It would focus
    on whether someone is anywhere within 3 miles, say, and ignore the
    differences between being 1, 2, or 3 miles away. Most proximity analysis has
    tended to look at this type of summary.

2.  Among the residents within X miles of the site, which groups live especially
    close to the facility? This question recognizes proximity is a continuous
    variable, and focuses on the difference between 1 mile, 1.5 miles, etc.
    However, it only looks at residents within the X miles radius area analyzed,
    so it fails to recognize that some groups tend to live more than 3 miles
    away, for example. This perspective does not take into account which groups
    are over-represented within the original total radius near a site.

Some functions like `distance_mean_by_group()` or `distance_by_group_by_site()` do
the second of these two types of analysis. They report, only among those
anywhere inside the radius, which groups are closer to the site.

In a specific location, for example, one residential population group could be
underrepresented within 3 miles, but those few who are in the group still might
live right next to the facility in which case their average distance would be
higher than that of any other group because this function only counts those
within the radius analyzed.

In some other location, the opposite could occur -- if one group is
over-represented within 3 miles, they still might all live in a community about
2.9 miles away from the site -- that would mean their distance from the site on
average is greater (or their proximity score is lower) than other groups within
3 miles of the site.

The question of whether to compare to Statewide or Nationwide or urban/rural or
other reference averages or percentiles is related to this question of how to
look at distances, or exposures or risk, just like it relates to how to look at
residential population group percentages. One could look at percentage rate within 1 mile, 2
miles, etc. all the way out until one was looking at the county overall, the
state overall, and eventually the nation overall. Selecting a single radius or
selecting a single reference area should be done with a recognition of what
questions one is actually trying to answer, and an understanding of how impacts
vary with distance from a particular type of facility or source of potential
risk.

If one is comparing residential population groups in terms of distance (or risk level), or
if one is comparing % at each distance (or risk level), the
implicit assumption is that there is some "expected" rate, and/or some
"equitable" or "proportionate" % or ratio or risk.

## CHOICE OF RADIUS AND UNCERTAINTY DUE TO A SMALL RADIUS WHERE BLOCKS ARE LARGE

Choosing a radius (or polygon) that is small relative to local Census blocks can
lead to significant uncertainty in EJAM estimates, so it is important
to understand the details if one wants to use a small radius especially in rural
(low population density) areas.

To help consider this uncertainty, EJAM reports how many block centroids were
found inside each area (inside a circular buffer defined by the selected radius,
or inside a polygon that is from a shapefile). That count of blocks is found in
a column of the spreadsheet output provided by the web app and also the table
called results_bysite that is one output of the ejamit() function.

You could also [Map all sites with popup at each saying how many blocks were
found nearby] and therefore might have more uncertainty in counts nearby.

```{r barplot_How_many_blocks_are_within_1_mile, fig.cap="How many blocks are within 1 mile of these 1,000 facilities?", fig.alt="barplot with count of facilities on y axis, 4 bars showing around 150 sites have <10 blocks within 1 mile, same for about 10-29 blocks nearby, around 300 sites have 30-100 blocks nearby, and >400 sites have over 100 blocks nearby"}
# out <- ejamit(testpoints_1000, radius = 1)
# out$results_bysite$blockcount_near_site
out <- testoutput_ejamit_1000pts_1miles

barplot(
  table(cut(
    out$results_bysite$blockcount_near_site,  
    c(-1, 9, 29, 100, 1000)
  )),
  names.arg = c("< 10 blocks", "10-29", "30-100", "> 100 blocks"),
  main = "How many blocks are within 1 mile of these 1,000 facilities?",
  ylab = "# of facilities", 
  xlab = "# of blocks nearby"
)
```

For more details about distance adjustments, overlaps of circles, etc.

This function prints a very large amount of diagnostic information, and provides
a barplot histogram showing in this case that almost none of the 1000 sites have
zero blocks within a mile but roughly 10-15% have under 10 blocks nearby and a
similar share have only 10-29 blocks nearby.

```{r diagnosticsdetails, eval=FALSE, echo=TRUE}
# (Printed information is lengthy)

  getblocks_diagnostics(
  testoutput_getblocksnearby_1000pts_1miles,
  # getblocksnearby(testpoints_1000, radius = 1, quiet = T),
  detailed = T, see_pctiles = T
  )
```

### Suggestions on radius and uncertainty

Here are some suggestions about how to consider the radius in relation to
uncertainty where blocks are large:

-   A closer look at uncertainty and care in communicating uncertainty may be
    needed where a circle or polygon contains fewer than about 30 block
    centroids. That is especially important if it contains fewer than about 10,
    and essential if it contains only 1 or zero block centroids.
-   Using a radius of 5 miles or more does not raise these issues in 99% of US
    locations where EPA-regulated facilities are found.
-   A radius of 3 miles might need a closer look for about 1% to 5% of typical
    sites in the US.
-   A radius of 1 mile or less requires caution and understanding of the issues
    at a significant share of locations in the US (about 1 in 4 locations might
    need a closer look to check for uncertainties).
-   A 0.5 mile radius should not be used without cautious interpretation or
    offline analysis in most locations where EPA-regulated facilities are
    located.
-   A 0.25 mile radius should only be used on a case-by-case basis where each
    location is examined individually and other methods are likely more suited
    for the analysis of those sites.

These considerations are explained further in the discussion below.

Residential population group counts and percentages or environmental indicators are calculated
from block group residential population data and environmental indicators and an estimate of
what fraction of each block group is inside each site. For proximity analysis
that means a circle is drawn around a point using a radius, and for shapefiles a
similar approach is used. In either case, the fraction of the block group
counted as inside the area analyzed is based on which block centroids (each is
technically called a block "internal point") are inside the circle or polygon.
All the residents of a block are assumed to be inside if the block centroid is
inside. This is exactly true unless a block is on the edge of the circle or
polygon. Even for the ones on the edge, some centroids are just outside and some
just inside the shape, so the contributions of some blocks are over-counted and
other under-counted, but those tend to cancel each other out in the sense that it
is unlikely they would all be under-counted, for example. Still, when a large
share of the block points in circle or polygon are from blocks not entirely
inside, uncertainty is higher than when the vast majority of blocks are entirely
inside. In other words, if the circle or polygon has a very large number of
blocks in it, uncertainty is lower because only a small fraction are along the
edge and bisected. If a radius of 3 miles is used, the area is 28 square miles.
If the blocks in that location are only about 0.28 square miles each, the circle
might contain or partly contain about 100 blocks.

The dataset used by EJAM called blockwts has a column called block_radius_miles
that is what the radius would be if the block were circular, and it was created
based on area = pi \* block_radius_miles\^2 or block_radius_miles = sqrt(area /
pi) where area is in square miles.

### Details on the blocks found near one site

#### Table of distances between each site and each block

Use `getblocksnearby()` to quickly find residents/blocks that are within a
specified distance, as a table of distances between sites and nearby blocks.

```{r getblocksnearby1, eval=TRUE, echo=TRUE}
sitepoints <- testpoints_10[1:2, ]

sites2blocks <- getblocksnearby(
  sitepoints = sitepoints,
  radius = 3.1
)
head(sites2blocks)
```

#### Detailed stats on blocks found near site(s)

```{r getblocks_diagnostics_simple, fig.cap="Example of getblocks_diagnostics() to see tables and histogram barplot of how many blocks are within 3.1 miles of these 2 sites", fig.alt="Example of getblocks_diagnostics() to see tables and histogram barplot of how many blocks are within 3.1 miles of these 2 sites, showing in this case they have at least 30 each", echo=TRUE, eval=TRUE}
x <- getblocks_diagnostics(sites2blocks)

# x <- getblocks_summarize_blocks_per_site(sites2blocks) 
# print(x) shows more info returned invisibly
```

#### Map 1 site to inspect the blocks nearby

Clicking on a block point provides a popup window showing information such as
this:

```         
blockfips: 131850102031056
blockid: 1788737
blocklat: 30.9913730000001
blocklon: -83.3753460999999
distance: 1.03614020347595
distance_unadjusted: 1.03614020347595
blockwt: 0
blockpop: 0
pop_nearby: 6237
bgpop: 1281
bgfips: 131850102031
bgid: 64286
ejam_uniq_id: 1
blockcount_near_site: 219
```

```{r plotblocksnearby2, fig.cap="Example of using plotblocksnearby() to see a map of the blockpoints near a site", eval=TRUE, echo=TRUE, fig.height=5, fig.width=5}
x <- plotblocksnearby(testpoints_10[1, ], radius = 3, returnmap = F)
# Set returnmap= TRUE to actually return a leaflet map
```

## POPULATION DENSITY -- WHY THE AVG SITE AND AVG RESIDENT ARE SO DIFFERENT

Reporting EJAM information summarized for the average site gives very different
answers than reporting on the average resident near any one or more of those
sites. The average site and average resident are completely different because
most of the residents live near just a few of the sites -- the ones with higher
population density -- when one is using a fixed radius at all sites, such as 3
miles from each site. Taking the average of sites gives equal weight to each
site, even the ones with very few residents around them. Taking the average of
all residents near all the sites gives equal weight to each person, so
conditions near certain sites affect more people and have more influence on that
average.

### Sites vary widely in count of blocks nearby, depending on population density (which is closely related to block area in square miles)

-   what blocks are near each site
-   how far are they
-   how many blocks are typically near a given site (population density varies)
-   how many sites are near a block (residents with \> 1 site nearby)

```{r popshare_at, eval=TRUE, echo=TRUE}
out <- testoutput_ejamit_100pts_1miles
cat("  ", popshare_p_lives_at_what_pct(out$results_bysite$pop, p = 0.50, astext = TRUE), "\n")
cat("  ", popshare_at_top_n(out$results_bysite$pop, c(1, 5, 10), astext = TRUE), "\n\n")

```

Find all blocks nearby each site

```{r getblocksnearby2, eval=TRUE, echo=TRUE}
radius <- 3
sitepoints <- testpoints_100
sites2blocks <- getblocksnearby(sitepoints, radius, quadtree = localtree, quiet = TRUE)
# testoutput_getblocksnearby_10pts_1miles is also available as an example
names(sites2blocks)
```

Very few blocks are within a radius of 1/4 mile.

Hundreds are often within 1 mile, but sometimes there are only a handful or even
zero.

```{r s2b_stats_xyz, eval=TRUE, echo=TRUE}
library(data.table)

s2b_stats <- sites2blocks[ , .(
  avgDistance = round(mean(distance), 2),
  blocksfound = .N, 
  blocks_within_1mile = sum(distance <= 1),
  blocks_within_0.75   = sum(distance <= 0.75),
  blocks_within_0.25  = sum(distance <= 0.25)
), by = 'ejam_uniq_id'][order(blocksfound), ]

setorder(s2b_stats, ejam_uniq_id)
head(s2b_stats)
```

### CDF of how many blocks are nearby a site

```{r plot_count_of_blocks_nearby, fig.cap="How many blocks are near each of these 100 facilities?", fig.alt="scatterplot of 100 ranked sites with y axis showing count of blocks nearby rising from zero to almost 800 blocks at a given site, and around 150-200 at the average site", eval=TRUE, echo=FALSE, fig.height=5, fig.width=6}

plot(sort(s2b_stats$blocks_within_1mile), 
     main = "How many blocks are near each facility?", 
     ylab = "# of blocks (whose internal point is) within 1 mile of each site", 
     xlab = paste0(nrow(s2b_stats)," facilities ranked by # of blocks nearby"))
abline(h = quantile(s2b_stats$blocks_within_1mile, probs = (0:4) * 0.25))
abline(h = mean(s2b_stats$blocks_within_1mile), col = "red")
```

### Histogram and table showing how many blocks are nearby a site

```{r histoblocks, eval=TRUE, echo=TRUE, fig.cap="Example of Histogram and table showing how many blocks are within 3 miles of a site", fig.alt="histogram with count of sites on y axis and how many blocks on x axis up to 6,000 showing huge share of sites have fewer than 500 blocks within 3 miles, but some have over 5,000 blocks within 3 miles of a single site.", fig.height=6, fig.width=7}

hist(sites2blocks[,.N, by = "ejam_uniq_id"][, N], 20, 
     xlab = "How many blocks are nearby?", 
     ylab = "Frequency (# of sites)", 
     main = "A given site may have zero to hundreds of blocks nearby", 
     sub = "A typical site in this example has about 100 blocks nearby")
```

```{r DTdatatable, eval=FALSE, echo=TRUE}
DT::datatable(s2b_stats,  rownames = FALSE)
# more summaries showing there may be only 1 block or hundreds within 1 mile
```

### Summary stats on how many blocks are within each radius

```{r quantiles_of_block_count_nearby}
# Just within 1 mile
summary(sites2blocks[distance <= 1, .N, by = "ejam_uniq_id"][, N])
# or
quantile(s2b_stats$blocks_within_1mile, probs = (0:4) * 0.25)

# Within each distance
summary(s2b_stats)
# t(summary(s2b_stats))
```

### Map all sites with popup at each saying how many blocks were found nearby

```{r s2b_stats_leaflet, eval=TRUE, echo=TRUE, fig.cap="Example of mapfast() for seeing how many blocks are at each site", fig.height=5, fig.width=5}
## done previously:
# radius <- 3
# sitepoints <- testpoints_100

out <- ejamit(sitepoints = sitepoints, 
              radius = radius, include_ejindexes = F)

few <- out$results_bysite$blockcount_near_site < 30

mapthis <- cbind(
  sitepoints, 
  out$results_bysite[, c(
    "pop", "bgcount_near_site", "blockcount_near_site"
    )],
  NOTE = ifelse(few, "< 30 blocks here", "")
  )

# Show in red the sites with very few blocks nearby, suggesting more uncertainty in residential population group counts

mm <- mapfast(mapthis, radius = radius, color = 'navy')
mm |> leaflet::addCircles(
    lng = mapthis$lon[few], 
    lat = mapthis$lat[few], 
    color = "red", radius = radius * 2 * meters_per_mile,
    popup = popup_from_any(mapthis[few, ])
)
```

#### Some places have very few -- if any -- blocks within 1 mile

```{r s2b_stats_more, eval=TRUE, echo=TRUE, fig.height=5, fig.width=5}
tail(s2b_stats[order(s2b_stats$blocks_within_1mile, decreasing = T), 
               c('ejam_uniq_id', 'blocks_within_1mile')], 3) 
```

#### Some places have hundreds nearby: a 1 mile radius is huge within a dense urban area

```{r s2b_stats2, eval=TRUE, echo=TRUE, fig.height=5, fig.width=5}
head(s2b_stats[order(s2b_stats$blocks_within_1mile, decreasing = T), 
               c('ejam_uniq_id', 'blocks_within_1mile')], 3)
```

```{r densest_plotblocks, eval=TRUE, echo=TRUE, fig.height=5, fig.width=5}
densest <- s2b_stats$ejam_uniq_id[order(
  s2b_stats$blocks_within_1mile, decreasing = T)][1]
leastdense <- s2b_stats$ejam_uniq_id[order(
    s2b_stats$blocks_within_1mile, decreasing = F)][1]
```

```{r, doplotblocksnotshowoutput, fig.cap="Map very high density area with 4,596 blocks near the site, via plotblocksnearby()", eval=TRUE, echo=FALSE}
# radius <- 3
sitepoints <- testpoints_100
plotblocksnearby(sitepoints = sitepoints[densest, ])
```

```{r showplotblockscommandnotdoit, eval=FALSE, echo=TRUE, include=TRUE}
plotblocksnearby(sitepoints = sitepoints[densest, ])
```

```{r, doplotblocksnotshowoutput_lowdensity, fig.cap="Map very low density area with 11 blocks near the site, via plotblocksnearby()", eval=TRUE, echo=FALSE}
plotblocksnearby(sitepoints = sitepoints[ leastdense, ])
```

```{r showplotblockscommandnotdoit_lowdensity, eval=FALSE, echo=TRUE, include=TRUE}
plotblocksnearby(sitepoints = sitepoints[ leastdense, ])
```

Within a 1 mile radius, the blocks found tend to be about 2/3 of a mile from the
site at the center.

```{r eval=TRUE, echo=TRUE}
summary(s2b_stats$avgDistance)
```
