---
title: "NAICS (Industries)"
description: "NAICS (Industries)"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{NAICS (Industries)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
---

```{r developernote, eval=FALSE, echo=FALSE, include=FALSE}
#  *>>>>>>>>>> Developer note: vignettes need to be tested/edited/rebuilt regularly <<<<<<<<<<<*
#    - **See ?pkgdown::build_site** and script in EJAM/data-raw/- EJAM uses the pkgdown R package to build help and articles/ vignettes as web pages
```

```{r SETUP_default_eval_or_not, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.height = 6, fig.width = 7,
  comment = "#>"
)
knitr::opts_chunk$set(eval = FALSE)
# https://r-pkgs.org/vignettes.html
```

```{r libraryEJAM, eval=TRUE, echo=FALSE, include=FALSE}
# rm(list = ls()); golem::detach_all_attached(); devtools::load_all()
 
if (!exists("blockgroupstats")) {library(EJAM)} # use installed version only if pkg not yet attached

dataload_dynamic('all') # varnames = all  currently means all defined by .arrow_ds_names

indexblocks()
```

# USING NAICS AND SIC CODES TO LOCATE FACILITIES BY INDUSTRY

EJAM helps select regulated sites based on industrial classification, using
NAICS or SIC code. Finding the right NAICS and finding all the right sites by
NAICS is complicated. Doing so requires understanding the NAICS system and the
FRS dataset, and the functions in EJAM that help find or use NAICS codes.

NAICS/SIC categories can be explored in a few ways:

-   [Key EJAM functions for using
    NAICS/SIC](../reference/index.html#naics-sic-datasets)
-   [NAICS.com website](https://www.naics.com) with extensive information [about
    NAICS](https://www.naics.com/everything-naics/) and
    [SIC](https://www.naics.com/everything-sic/)
-   EPA [FRS Facility Industrial Classification Search
    tool](https://www.epa.gov/frs/frs-query#industrial) where you can find
    facilities based on NAICS or SIC.
-   EPA APIs exist that can be used for similar queries.

Some key functions include `regid_from_naics()`, `latlon_from_naics()`,
`frs_from_naics()`, `naics_findwebscrape()`, and `naics_categories()`. These
functions can help find EPA FRS sites based on naics codes or titles. They rely
on `frs_by_naics` (a data.table), and `naics_from_any()` for querying by code or
title of category.

Files and dataset examples related to NAICS:

```{r}
topic = "naics"
cbind(data.in.package  = sort(grep(topic, EJAM:::datapack()$Item, value = T)))
cbind(files.in.package = sort(basename(testdata(topic, quiet = T))))

```

**Important points:**

-   Note that a very large fraction of all FRS sites (as obtained for use in
    EJAM) lack NAICS code!

-   Note that EJAM may query FRS sites differently than the FRS search tool or
    other query tools would.

-   Note that (NAICS.com) reports many more businesses for a given 6-digit
    category than the FRS shows, which might be due to FRS only including
    EPA-regulated sites but also due to data gaps.

-   Note the difference between `children = TRUE` and `children = FALSE` in EJAM
    functions like `latlon_from_naics()`

-   Note that searching on a 6-digit code misses parent categories you may want.
    The FRS data on NAICS by site is inconsistent in how many digits are
    reported for the NAICS (explained below).

A given site might be listed in the FRS as being under one or more NAICS codes
of various lengths, such as only a parent code (large grouping), only a detailed
code (6-digit), or some combination of codes and their subcategories.

And the same title, like "Petroleum Refineries," may be assigned by the NAICS
system to the category but also a subcategory, as with codes 32411 and 324110.
The function `naics_from_any()` shows what codes and title exist in the NAICS
system.

Also, certain terms appear in the online description of a NAICS but not in the
title of the NAICS -- the function `naics_findwebscrape()` helps with those
cases, e.g., compare these:

```{r webscrape_vs, eval=FALSE}
naics_findwebscrape("cement")

naics_from_any("cement")
```

Compare also these:

```{r webscrape_vs2, eval=FALSE}
naics_findwebscrape("refiner")
# reports "324110" (Petroleum Refineries) and other related industries, but not the 5-digit "32411" (also Petroleum
Refineries).

naics_from_any("refiner")
# reports "324110" and "32411" but not other related industries.
```

Using `naics_findwebscrape()` finds only the 6-digit codes that match on title
or description, so it would find some codes not found by `naics_from_any()`
which does not query description, but could lead to missing some facilities in
the sense that the 6-digit code does not cover the sites listed in FRS under
only the 5-digit code for Petroleum Refineries (not the 6-digit).

It is important to note that searching on a 6-digit code misses parent
categories that may include sites you expect to find:

`frs_from_naics()` used as `frs_from_naics("324110", children = F)[,1:5]` finds a few hundred sites, but it
fails to find some sites you would find using
`frs_from_naics()` used as `frs_from_naics("32411", children = F)[,1:5]`

The code example below shows that the FRS dataset has some facilities listed
under the 5-digit "32411" code only, some with the 6-digit "324110" code only,
and some with both codes:

```{r 5vs6digitNAICS_SHOWNOTDO, eval=FALSE, include=TRUE}
hasboth = intersect(
frs_from_naics("32411",  children = F)[,1:5]$REGISTRY_ID,
frs_from_naics("324110", children = F)[,1:5]$REGISTRY_ID
)
hasonly6digit = setdiff(
frs_from_naics("32411",  children = F)[,1:5]$REGISTRY_ID,
frs_from_naics("324110", children = F)[,1:5]$REGISTRY_ID
)
hasonly5digit = setdiff(
frs_from_naics("324110", children = F)[,1:5]$REGISTRY_ID,
frs_from_naics("32411",  children = F)[,1:5]$REGISTRY_ID
)
```

```{r 5vs6digitNAICS_DONOTSHOW, eval=TRUE, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
## include=F etc. avoids showing frs_arrow is loading from...

hasboth = intersect(
frs_from_naics("32411",  children = F)[,1:5]$REGISTRY_ID,
frs_from_naics("324110", children = F)[,1:5]$REGISTRY_ID
)
hasonly6digit = setdiff(
frs_from_naics("32411",  children = F)[,1:5]$REGISTRY_ID,
frs_from_naics("324110", children = F)[,1:5]$REGISTRY_ID
)
hasonly5digit = setdiff(
frs_from_naics("324110", children = F)[,1:5]$REGISTRY_ID,
frs_from_naics("32411",  children = F)[,1:5]$REGISTRY_ID
)
```


```{r 5vs6digitNAICS_results, eval=TRUE, message=FALSE}

length(hasonly5digit)  # Most of the FRS sites here
length(hasonly6digit)
length(hasboth)
```

## Examples of some NAICS/SIC functions

```{r naics_examples, eval=FALSE}
naics_from_any(naics_categories(3))[order(name),.(name,code)][1:10,]
naics_from_any(naics_categories(3))[order(code),.(code,name)][1:10,]

naics_from_code(211)
naicstable[code==211,]
naics_subcodes_from_code(211)
naics_from_code(211,  children = TRUE)
naicstable[n3==211,]
NAICS[211][1:3] # wrong
NAICS[NAICS == 211]
NAICS["211 - Oil and Gas Extraction"]

naics_from_any("plastics and rubber")[,.(name,code)]
naics_from_any(326)
naics_from_any(326, children = T)[,.(code,name)]
naics_from_any("plastics", children=T)[,unique(n3)]
naics_from_any("pig")
naics_from_any("pig ") # space after g

# naics_from_any("copper smelting")
# naics_from_any("copper smelting", website_scrape=TRUE)
# browseURL(naics_from_any("copper smelting", website_url=TRUE) )

a = naics_from_any("plastics")
b = naics_from_any("rubber")
fintersect(a,b)[,.(name,code)] #  a AND b
funion(a,b)[,.(name,code)]     #  a OR  b
naics_subcodes_from_code(funion(a,b)[,code])[,.(name,code)]   #  plus children
naics_from_any(funion(a,b)[,code], children=T)[,.(name,code)] #  same

NROW(naics_from_any(325))
#[1] 1
NROW(naics_from_any(325, children = T))
#[1] 54
NROW(naics_from_any("chem"))
#[1] 20
NROW(naics_from_any("chem", children = T))
# [1] 104
```
